Divergent morphology of elliptical bifaces from sites in the American Southeast articulates with raw material differences

### 2022-08-05 supp/index.Rmd
<edit index>

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, those bifaces from Jowell Farm are assumed to have been manufactured using Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape at the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests both a shared mental template, and a size differential is potentially driven by raw material. Additionally, elliptical biface size in the large size class does not differ between raw materials; however, the small size class does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening trajectory, where bifaces from each site/raw material enter their uselife at a similar shape and size, diverging significantly in size---but not shape---throughout that uselife. In following with our expectations, greater shape diversity was found to occur in the small size class, where those bifaces from Jowell Farm were more diverse than those from the Millsap Cache. Discrete raw material differences are posited as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area.

>>>>

Elliptical bifaces from the Millsap Cache and Jowell Farm (41AN13) sites comprise two of the largest samples of this ill-understood lithic tool from the ancestral Caddo area. Similar to Gahagan bifaces, elliptical bifaces from Jowell Farm are assumed to have been manufactured from Edwards chert---due to their size, the colours that they fluoresce beneath low and high frequency UV light, and the paucity of similarly large locally-available raw material sources---and acquired by the Caddo through trade with central Texas groups. Bifaces from the Millsap Cache are thought to have been manufactured from Kay County flint, providing for an empirical assessment of whether elliptical biface morphology differs as a function of raw material. The sample from each site was subsequently divided into two size classes; one reflective of initial production/manufacture (larger), and the other associated with resharpening and/or retouch (smaller). Size classes were used to assess whether modifications by Caddo knappers may have yielded a similar---convergent---knife shape in the smaller size class, and to test the hypothesis that greater morphological variation would be apparent in the small size class due to the variable idiosyncracies related to resharpening and/or retouch. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential driven by raw material differences. Additionally, size does not differ between raw materials in the large size class; however, in the small size class, it does. This finding aids in advancing the argument that elliptical biface morphology diverges through the resharpening lifecycle, where bifaces manufactured using both raw materials enter their uselife at a similar shape and size, but diverge significantly in size---but not shape---throughout that uselife. In following with our expectations, greater shape diversity was found to occur in the small size class, where bifaces from Jowell Farm are more diverse than those from the Millsap Cache. Discrete raw material differences are advanced as the drivers of the extant morphological differences found in elliptical bifaces from the ancestral Caddo area in the American Southeast.
